---
title: "Problem Set 5"
author: "Nikita Yadav"
date: "Feb 18, 2023"
output: html_document
---

### Load and Check Data

```{r}
library(readr)
library(tidyverse)
library(ggplot2)
gapminder_data <- read_delim("gapminder.csv")
```
This dataset has `r nrow(gapminder_data)` rows and `r ncol(gapminder_data)`
columns.
```{r}
print(head(gapminder_data))
```
My small sample of data does look ok! It does look like there are a lot of missing values though.

### Descriptive Statistics

```{r}
num_iso3 <- length(unique(gapminder_data$iso3))
num_name <- length(unique(gapminder_data$name))
num_iso2 <- length(unique(gapminder_data$iso2))
```
Each of the three variables to be evaluated for question 1 in this section has a different number of unique elements. According to `iso3`, there are `r num_iso3` countries. According to `name`, there are `r num_name` countries. And according to `iso2`, there are `r num_iso2` countries.

iterate through num_names in iso2_only dataframe and if any value is greater than 1 pull out the iso2 code and save it as a vector. find where iso2 == NA and print gapminder_data$name that corresponds. get index
```{r}
iso2_only <- gapminder_data %>%
  group_by(iso2) %>%
  summarize(num_names = n_distinct(name))

iso2_only$iso2[which(iso2_only$num_names > 1)]

iso2_is_na <- is.na(gapminder_data$iso2)
missing_iso2 <- c()
for(i in which(iso2_is_na == TRUE)) {
  missing_iso2<- c(missing_iso2, gapminder_data$name[i])
}
missing <- unique(missing_iso2)

```
There aren't iso2 codes that correspond to more than one name. There are two values in the column `name` in the dataframe that don't have an `iso2` code. They are `r missing[1]` and `r missing[2]`.

```{r}
name_only <- gapminder_data %>%
  group_by(name) %>%
  summarize(num_iso3 = n_distinct(iso3))

name_only$num_iso3[which(name_only$num_iso3 > 1)]

which(name_only$num_iso3 > 1)

name_only$name[250]

na_country<- gapminder_data %>%
  select(iso3, name)%>%
  filter(is.na(name))%>%
  distinct(iso3)

```
There are 4 `iso3` codes that don't correspond to a country name. Instead they correspond to an `NA` value. These codes are `r paste(na_country$iso3, collapse = ", ")`.

```{r}
min_year <- min(gapminder_data$time, na.rm=TRUE)
max_year <- max(gapminder_data$time, na.rm=TRUE)
```

The minimum year in this dataset is `r min_year` and the maximum year is `r max_year`.

### CO2 Emissions

```{r}
missing_co2 <- gapminder_data %>%
  group_by(time) %>%
  summarize(num_na_co2 = sum(is.na(co2)),
            num_na_co2_PC = sum(is.na(co2_PC)))

max_missing_value <- max(missing_co2$num_na_co2)
year_indices <- which(missing_co2$num_na_co2==max_missing_value)

missing_years <- c()
for(i in year_indices) {
  missing_years <- c(missing_years, missing_co2$time[i])
}
```
The years with the most missing CO2 data are `r paste(missing_years, collapse = ", ")` They each have `r max_missing_value` missings.

```{r}
countries <- c("United States of America", "China", "India", "Afghanistan", "Sweden")
selected_countries <- gapminder_data %>%
  filter(name %in% countries)

ggplot(data = selected_countries) +
  geom_line(mapping = aes(x = selected_countries$time, 
                          y = selected_countries$co2,
                          color = selected_countries$name)) +
  labs(
    title = "Total CO2 Emissions Over Time",
    x = "Year",
    y = "CO2 Emissions (kt)",
    color = "Country Name"
  )
```

Afghanistan and Sweden have the lowest CO2 emissions, both lines are basically at the 0 mark. The US, China, and India have slowly risen in emissions over the years, with China emitting the most in recent years.

```{r}
ggplot(data = selected_countries) +
  geom_line(mapping = aes(x = selected_countries$time,
                          y = selected_countries$co2_PC,
                          color = selected_countries$name)) +
  labs(
    title = "CO2 Emissions Per Capita over Time",
    x = "Year",
    y = "CO2 Emissions Per Capita (metric tons)",
    color = "Country"
  )
```

This plot is a bit surprising in comparison to the previous one. Afghanistan still has the least amount of CO2 emissions, but this time India is accompanying Afghanistan along the x axis of the plot. The US has the most metric tons of CO2 emissions per capita, far above any of the other countries. 

```{r}
avg_co2 <- gapminder_data %>%
  group_by(region, time, .groups="rowwise") %>%
  summarize(
    avg_co2_pc = mean(co2_PC, na.rm=TRUE)
  )
```

This code chunk produced a new data frame that is grouped by the region and time columns in the original data frame. I added an additional column with the average CO2 emissions per capita for each region, per year. 

```{r}
time_filter <- c("1960", "2016")
avg_co2 <- na.omit(avg_co2)%>%
  filter(time %in% c("1960", "2016"))

ggplot(avg_co2, aes(avg_co2$region, avg_co2$avg_co2_pc,
                    fill=as.factor(avg_co2$time)))+
  geom_bar(stat="identity", position="dodge") +
  labs(
    title = "Avg CO2 Emissions Per Region (1960 and 2016)",
    x = "Region",
    y = "Avg CO2 Emissions per capita") +
  scale_fill_manual(
    values = c("#e41a1c", "#377eb8"),
    name="Year"
  )
```

The code chunk above produces a barplot that visualizes the average CO2 emissions per capita in each unique region from the original dataset for the years: `r paste(time_filter, collapse = ", ")`.

group original df by region and time

```{r}
checking <- gapminder_data %>%
  group_by(time, co2_PC) %>%
  filter(time==2019)

CO2_2016 <- gapminder_data %>%
  drop_na(co2_PC) %>%
  filter(time==2016) %>%
  group_by(region, time, name, .groups="rowwise") %>%
  summarize(avg_co2_pc = mean(co2_PC))

largest_emitters <- CO2_2016 %>%
  group_by(region) %>%
  arrange(desc(avg_co2_pc)) %>%
  slice(1:3) %>%
  select(region, name, avg_co2_pc)

smallest_emitters <- CO2_2016 %>%
  group_by(region) %>%
  arrange(avg_co2_pc) %>%
  slice(1:3) %>%
  select(region, name, avg_co2_pc)
```
Because 2019 is missing data for the co2_PC variable, I did this problem using 2016 stats. The  largest CO2 emitters per region are `r paste(largest_emitters$name, collapse=", ")`. The smallest CO2 emiiters per region are `r paste(smallest_emitters$name, collapse=", ")`.

### GDP Per Capita