---
title: "Problem Set 5"
author: "Nikita Yadav"
date: "Feb 18, 2023"
output: html_document
---

### Load and Check Data

```{r}
library(readr)
library(tidyverse)
library(ggplot2)
gapminder_data <- read_delim("gapminder.csv")
```
This dataset has `r nrow(gapminder_data)` rows and `r ncol(gapminder_data)`
columns.
```{r}
print(head(gapminder_data))
```
My small sample of data does look ok! It does look like there are a lot of missing values though.

### Descriptive Statistics

```{r}
num_iso3 <- length(unique(gapminder_data$iso3))
num_name <- length(unique(gapminder_data$name))
num_iso2 <- length(unique(gapminder_data$iso2))
```
Each of the three variables to be evaluated for question 1 in this section has a different number of unique elements. According to `iso3`, there are `r num_iso3` countries. According to `name`, there are `r num_name` countries. And according to `iso2`, there are `r num_iso2` countries.

```{r}
iso2_only <- gapminder_data %>%
  group_by(iso2) %>%
  summarize(num_names = n_distinct(name))

iso2_only$iso2[which(iso2_only$num_names > 1)]

iso2_is_na <- is.na(gapminder_data$iso2)
missing_iso2 <- c()
for(i in which(iso2_is_na == TRUE)) {
  missing_iso2<- c(missing_iso2, gapminder_data$name[i])
}
missing <- unique(missing_iso2)

```
There aren't iso2 codes that correspond to more than one name. There are two values in the column `name` in the dataframe that don't have an `iso2` code. They are `r missing[1]` and `r missing[2]`.

```{r}
name_only <- gapminder_data %>%
  group_by(name) %>%
  summarize(num_iso3 = n_distinct(iso3))

name_only$num_iso3[which(name_only$num_iso3 > 1)]

which(name_only$num_iso3 > 1)

name_only$name[250]

na_country<- gapminder_data %>%
  select(iso3, name)%>%
  filter(is.na(name))%>%
  distinct(iso3)

```
There are 4 `iso3` codes that don't correspond to a country name. Instead they correspond to an `NA` value. These codes are `r paste(na_country$iso3, collapse = ", ")`.

```{r}
min_year <- min(gapminder_data$time, na.rm=TRUE)
max_year <- max(gapminder_data$time, na.rm=TRUE)
```

The minimum year in this dataset is `r min_year` and the maximum year is `r max_year`.

### CO2 Emissions

```{r}
missing_co2 <- gapminder_data %>%
  group_by(time) %>%
  summarize(num_na_co2 = sum(is.na(co2)),
            num_na_co2_PC = sum(is.na(co2_PC)))

max_missing_value <- max(missing_co2$num_na_co2)
year_indices <- which(missing_co2$num_na_co2==max_missing_value)

missing_years <- c()
for(i in year_indices) {
  missing_years <- c(missing_years, missing_co2$time[i])
}
```
The years with the most missing CO2 data are `r paste(missing_years, collapse = ", ")` They each have `r max_missing_value` missings.

```{r}
countries <- c("United States of America", "China", "India", "Afghanistan", "Sweden")
selected_countries <- gapminder_data %>%
  filter(name %in% countries)

ggplot(data = selected_countries) +
  geom_line(mapping = aes(x = selected_countries$time, 
                          y = selected_countries$co2,
                          color = selected_countries$name)) +
  labs(
    title = "Total CO2 Emissions Over Time",
    x = "Year",
    y = "CO2 Emissions (kt)",
    color = "Country Name"
  )
```

Afghanistan and Sweden have the lowest CO2 emissions, both lines are basically at the 0 mark. The US, China, and India have slowly risen in emissions over the years, with China emitting the most in recent years.

```{r}
ggplot(data = selected_countries) +
  geom_line(mapping = aes(x = selected_countries$time,
                          y = selected_countries$co2_PC,
                          color = selected_countries$name)) +
  labs(
    title = "CO2 Emissions Per Capita over Time",
    x = "Year",
    y = "CO2 Emissions Per Capita (metric tons)",
    color = "Country"
  )
```

This plot is a bit surprising in comparison to the previous one. Afghanistan still has the least amount of CO2 emissions, but this time India is accompanying Afghanistan along the x axis of the plot. The US has the most metric tons of CO2 emissions per capita, far above any of the other countries. 

```{r}
avg_co2 <- gapminder_data %>%
  group_by(region, time, .groups="rowwise") %>%
  summarize(
    avg_co2_pc = mean(co2_PC, na.rm=TRUE)
  )
```

This code chunk produced a new data frame that is grouped by the region and time columns in the original data frame. I added an additional column with the average CO2 emissions per capita for each region, per year. 

```{r}
time_filter <- c("1960", "2016")
avg_co2 <- na.omit(avg_co2)%>%
  filter(time %in% c("1960", "2016"))

ggplot(avg_co2, aes(avg_co2$region, avg_co2$avg_co2_pc,
                    fill=as.factor(avg_co2$time)))+
  geom_bar(stat="identity", position="dodge") +
  labs(
    title = "Avg CO2 Emissions Per Region (1960 and 2016)",
    x = "Region",
    y = "Avg CO2 Emissions per capita") +
  scale_fill_manual(
    values = c("#e41a1d", "#377eb9"),
    name="Year"
  )
```

The code chunk above produces a barplot that visualizes the average CO2 emissions per capita in each unique region from the original dataset for the years: `r paste(time_filter, collapse = ", ")`.

group original df by region and time

```{r}
checking <- gapminder_data %>%
  group_by(time, co2_PC) %>%
  filter(time==2019)

CO2_2016 <- gapminder_data %>%
  drop_na(co2_PC) %>%
  filter(time==2016) %>%
  group_by(region, time, name, .groups="rowwise") %>%
  summarize(avg_co2_pc = mean(co2_PC))

largest_emitters <- CO2_2016 %>%
  group_by(region) %>%
  arrange(desc(avg_co2_pc)) %>%
  slice(1:3) %>%
  select(region, name, avg_co2_pc)

smallest_emitters <- CO2_2016 %>%
  group_by(region) %>%
  arrange(avg_co2_pc) %>%
  slice(1:3) %>%
  select(region, name, avg_co2_pc)
```
Because 2019 is missing data for the co2_PC variable, I did this problem using 2016 stats. The  largest CO2 emitters per region are `r paste(largest_emitters$name, collapse=", ")`. The smallest CO2 emiiters per region are `r paste(smallest_emitters$name, collapse=", ")`.

### GDP Per Capita

```{r}
GDP_1960 <- gapminder_data %>%
  filter(time==1960)

ggplot(GDP_1960,
       aes(GDP_1960$GDP_PC, GDP_1960$lifeExpectancy,
           size = GDP_1960$totalPopulation,
           color = GDP_1960$region)) +
  geom_point() +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy",
    title = "GDP per capita vs life expectancy (1960)") +
  scale_color_brewer(palette = "Set1", name = "Region") +
  scale_size_continuous(range = c(1, 10), name = "Population Size")
```

Based on this plot, it looks like Africa and Asia have both the lowest GDP per capita and are on the lower end of the life expectancy spectrum. Europe comes up on top for life expectancy and GDP. The Americas and Oceania are in the middle of the two extremes. 

```{r}
GDP_2019 <- gapminder_data %>%
  filter(time==2019)

ggplot(GDP_2019,
       aes(GDP_2019$GDP_PC, GDP_2019$lifeExpectancy,
           size = GDP_2019$totalPopulation,
           color = GDP_2019$region)) +
  geom_point() +
  labs(
    x = "GDP per capita",
    y = "Life Expectancy",
    title = "GDP per capita vs life expectancy (2019)") +
  scale_color_brewer(palette = "Set1", name = "Region") +
  scale_size_continuous(name = "Population Size")

```

The range for GDP per capita values tripled between 1960 and 2019. The distribution for the most part looks similar to 1960, but life expectancy has increased a bit across all regions. Africa is still on the lowest end of both GDP and life expectancy and Europe is on the highest end for both variables. 

```{r}
time_filter <- c("1960", "2019")
avg_LE <- gapminder_data %>%
  filter(time %in% time_filter) %>%
  group_by(region, time, .groups="rowwise") %>%
  summarize(avg_LE = mean(lifeExpectancy, na.rm=TRUE))

head(avg_LE)
```

I included the first few rows from my new dataframe displaying the average life expectancy for each region for the years 1960 and 2019. The results do match the figures. For example, in 2019, Africa's average life expectancy is about 64 years and in the 2019 GDP plot, the center of the Africa points is right around the 64 or 65 year mark.

```{r}
LE_growth <- gapminder_data %>%
  filter(time %in% time_filter) %>%
  drop_na(lifeExpectancy)%>%
  group_by(region) %>%
  summarize(avg_le_growth = mean(lifeExpectancy[time == 2019] -
                                   lifeExpectancy[time == 1960]))

LE_growth %>%
  arrange(avg_le_growth)
```

Europe has had the smallest amount of growth in life expectancy from 1960 to 2019 and Asia had the highest amount of average growth between the two years. 

```{r}
gdp_hist <- gapminder_data %>%
  filter(time %in% time_filter) %>%
  drop_na(GDP_PC) %>%
  select(time, GDP_PC)

ggplot(data = gdp_hist) +
  geom_histogram(aes(x = gdp_hist$GDP_PC, fill=factor(time)), 
                 position = "dodge") +
  labs(x = "GDP per Capita",
       title = "GDP per capita for 1960 and 2019"
  )
```

```{r}
US_rank <- gapminder_data %>%
  filter(time %in% time_filter) %>%
  select(name, time, lifeExpectancy) %>%
  arrange(desc(lifeExpectancy))

le_rank_1960 <- US_rank %>%
  filter(time==1960) %>%
  drop_na(name) %>%
  #arrange(desc(lifeExpectancy))
  mutate(ranking = rank(desc(lifeExpectancy)))

le_rank_2019 <- US_rank %>%
  filter(time==2019) %>%
  drop_na(name) %>%
  mutate(ranking = rank(desc(lifeExpectancy)))

```

The ranking of the U.S. in terms of life expectancy in 1960 was `r paste(which(le_rank_1960$name=="United States of America"))` and in 2019 it was `r paste(which(le_rank_2019$name=="United States of America"))`.

```{r}
num_countries <- gapminder_data %>%
  filter(!is.na(lifeExpectancy)) %>%
  group_by(time) %>%
  summarize(n = n())

yearly_rank <- gapminder_data %>%
  filter(!is.na(lifeExpectancy)) %>%
  filter(!is.na(name)) %>%
  group_by(time) %>%
  mutate(ranking = rank(desc(lifeExpectancy))) %>%
  filter(name == "United States of America") %>%
  select(time, ranking)

relative_rank <- num_countries %>%
  left_join(yearly_rank, by = "time") %>%
  mutate(relative_rank = ranking / n)

head(relative_rank)
```

I included just the first few rows of my relative rank dataframe. With the relative rank, it is clear that the U.S. dropped in the rankings because of the increase in number of countries. Still, the U.S. is doing worse in terms of life expectancy than before.

I spent about 12 hours on this problem set!